---
title: "R for Reproducibility"
subtitle: "R Packages for Reproducibile Workflow"
author: "Daniela Palleschi"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    html-math-method: katex
    css: styles.css
    number-sections: true
    number-depth: 3
    code-overflow: wrap
    self-contained: true
  pdf:
    output-file: packages.pdf
    toc: true
    toc-depth: 2
    number-sections: true
    code-link: true
    code-overflow: wrap
    code-tools: true
    self-contained: true
editor: source
editor_options: 
  chunk_output_type: console
---

This is an overview of R packages that I use to maintain a reproducible workflow.

```{r, echo = F}
# set global options
knitr::opts_chunk$set(eval = F, # run each code chunk?
                      echo = T, # print code chunk?
                      message = F, # print messages (e.g., warnings)?
                      error = F, # print errors?
                      cache = T # cache? (Session > Restart R to clear cache)
                      ) 
```

# `here` package

The `here` package uses file path relative to the current working directory [@here-package]. If you're working within a RProject (please do so!) then this working directory is the folder containing the `.RProj` file.

```{r}
# install here package
install.packages("here")

# read in a csv file
readr::read_csv(here::here("folder", "subfolder", "subsubfolder", "file.csv"))
# or
readr::read_csv(here::here("folder/subfolder/subsubfolder/file.csv"))
```


# `pacman` package

The `pacman` package was designed to facilitate loading and installing packages [@pacman-package]. I like to use the `p_load()` function, which checks if listed packages are already installed. If so, it loads them (as with `library()`). If not, it installs and then loads them (as long as they're CRAN packages!). This facilitates sharing your code across machines or package libraries.
  
```{r}
# install pacman
install.packages("pacman")

# install and/or load packages
pacman::p_load(here,
               renv,
               kableExtra)
```

# `rbbt` package

The `rbbt` package contains an RStudio Addin for better integration of RStudio, Better Bibtex, and Zotero [@rbbt-package]. It can be installed via GitHub:

```{r}
#| eval: false
# install rbbt; remotes must first be installed
remotes::install_github("paleolimbot/rbbt")
```

You can follow this [blog post](https://gsverhoeven.github.io/post/zotero-rmarkdown-csl/) from Gertjan Verhoeven (May 2021) on integrating RStudio and Zotero for writing in Rmarkdown, assuming only that you have RStudio installed.

Caveat: currently some `rbbt` functions conflict with Quarto's cross-references, which also begin with `@`. For Quarto documents, I mainly use the `rbbt` Addin to insert Zotero citations while writing using a keyboard shortcut (see the blogpost for tips on how to set this up).


# `renv` package

The `renv` packages stores your project-local environment [@renv-package]. That is, it creates a time capsule of the R and package versions that you use *within an R Project*.

The benefits:

- your output is not dependent on whichever project version you currently have stored on your machine
  + e.g., if you come back to an analysis after a long time, the output should still be the same even if a certain package version has deprecated a function your analysis uses.
- makes the project *reproducible* for collaborators as well, who will likely have some different package versions (or will have missing packages) on their machine

The hard-to-get-your-head-around (imo):

- with each new R Project with a renv.lock file, you need to install all the necessary packages again (even if they're on your machine)
  + this is because each `renv` doesn't look beyond your project folder!
- remembering to update the `renv.lock` file frequently!

::: {.callout-warning}
## Running `renv` functions

Remember to always run `renv` functions ***in the console***! You do not want e.g., `renv` initialising a new `renv.lock` file every time you render your documents.

:::

## Activate or init

- to initialise a new *project-local environment*:

```{r}
renv::init()
```


## Snapshot

- update `renv.lock` file

```{r}
renv::snapshot()
```

After updating the `renv.lock` file, remember to commit/push these changes to git (if you're using it)!

::: {.callout-tip}
## Version control

After updating the `renv.lock` file (i.e., running `renv::snapshot()`), remember to commit/push these changes to git (if you're using git)! Otherwise, your `renv.lock` file will be outdated compared to your output.

:::

## Restore

- will restore your project to the most recent `renv.lock` file versions
  - this step follows `snapshot()`, which updates the `renv.lock` file
- very useful after updating R!

```{r}
renv::restore()
```

## Upgrade

- to update the `renv` package:

```{r}
# upgrade renv version
renv::upgrade()
```

## Update and Hydrate

```{r}
# updates all packages (stored in renv.lock) in the renv cache
renv::hydrate(update = "all")

# update should have no effect now, but doesn't hurt to check
renv::update()

# now take a snapshot with the updated packages
renv::snapshot()
```

::: {.callout-tip}
## Version control (repeated)

After updating the `renv.lock` file (i.e., running `renv::snapshot()`), remember to commit/push these changes to git (if you're using git)! Otherwise, your `renv.lock` file will be outdated compared to your output.

:::


# `usethis` package

The `usethis` package automates project set-up and development, and is also useful when creating a package (see the package [Documentation](https://usethis.r-lib.org/index.html) for more info) [@usethis-package].

```{r}
#| echo: false
#| eval: true
pacman::p_load(
  tidyverse,
  knitr,
  kableExtra
)
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-usethis
#| tbl-cap: "Summary of useful commands from the usethis package"

dplyr::tribble(
  ~"command", ~"function",
  "git_token_help()", "print crediential situation",
  "git_sitrep()", "print more verbose credential information",
  "git_vaccinate()", "add appropriate files to .gitignore (remember to stage, commit, and push)",
  "browse_github()", "view project GitHub",
  "browse_project() / _package()", "visit project-related pages",
  "browse_github_actions()", "view overview of GitHub actions",
  "git_credentials()", "print credentials",
  "create_github_token()", "create a Personal Access Token (PAT)",
  "gitcreds_set()", "add your PAT"
) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling()
```

# `gitcreds` package

Has some similar functionality to `usethis` with regards to using git/GitHub in an R project [@gitcreds-package]. See the package [GitHub page](https://github.com/r-lib/gitcreds) for more information.

```{r}
#| echo: false
#| eval: true
#| label: tbl-gitcreds
#| tbl-cap: "Summary of useful commands from the gitcreds package"

dplyr::tribble(
  ~"command", ~"function",
  "gitcreds_get()", "print git/GitHub credentials",
  "gitcreds_set()", "set/replace credentials",
  "gitcreds_delete()", "delete credentials"
) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling()
```

