---
title: "Troubleshooting"
author: "Daniela Palleschi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number-sections: true
---

```{r}
knitr::opts_chunk$set(eval = F, # change this to 'eval = T' to reproduce the analyses; make sure to comment out 
                      echo = T, # 'print code chunk?'
                      message = F, # 'print messages (e.g., warnings)?'
                      error = F,
                      warning = F)
```

In this document I record errors that I encounter and how I resolved them. Solutions will typically be from Googling, relevant links will be provided.


# Rmarkdown knitting PDF

Error message:

```{r, echo = T, eval = F}
tlmgr update --all --self
fmtutil [ERROR]: format directory not writable: /usr/local/texlive/2019/texmf-var/web2c
tlmgr: package repository https://mirror.clientvps.com/CTAN/systems/texlive/tlnet (not verified: gpg unavailable)
tlmgr install: package already present: l3kernel
! LaTeX Error: Mismatched LaTeX support files detected.
(LaTeX)        Loading 'expl3.sty' aborted!
(LaTeX)        
(LaTeX)        The L3 programming layer in the LaTeX format
(LaTeX)        is dated 2023-02-01, but in your TeX tree the files require
(LaTeX)        at least 2023-02-02.

Error: LaTeX failed to compile 03-exp123.tex. See https://yihui.org/tinytex/r/#debugging for debugging tips. See 03-exp123.log for more info.
In addition: Warning messages:
1: Removed 156 rows containing non-finite values (stat_ydensity). 
2: Removed 156 rows containing non-finite values (stat_boxplot). 
3: Removed 156 rows containing non-finite values (stat_summary). 
4: Removed 156 rows containing non-finite values (stat_summary). 
5: Removed 156 rows containing non-finite values (stat_ydensity). 
6: Removed 156 rows containing non-finite values (stat_boxplot). 
7: Removed 156 rows containing non-finite values (stat_summary). 
8: Removed 156 rows containing non-finite values (stat_summary). 
Execution halted
```

In the Terminal, I type

```{r}
kpsewhich --all --engine=pdftex pdflatex.fmt
```

And get:

```{r}
(base) administrators-MacBook-3:Lifetime Project danielapalleschi$ kpsewhich --all --engine=pdftex pdflatex.fmt
/usr/local/texlive/2019/texmf-var/web2c/pdftex/pdflatex.fmt
```

So my pdftex version is from 2019 or soemthing.

In the Terminal, I typed:

```{r}
sudo fmtutil-sys --all
```

And entered my laptop's administrator Password when prompted (it won't show that you're typing, but just type the password and hit Enter). Then Terminal was running for a couple of minutes, and ended with the output:

```{r}
Transcript written on mex.log.
fmtutil [INFO]: /usr/local/texlive/2019/texmf-var/web2c/pdftex/mex.fmt installed.
fmtutil [INFO]: Disabled formats: 6
fmtutil [INFO]: Successfully rebuilt formats: 45
fmtutil [INFO]: Total formats: 51
fmtutil [INFO]: exiting with status 0
```

When I went to knit a PDF I still got the same errors.

Then I ran 

```{r}
fmtutil-user --all
```

No password required, it ran again for a few minutes. At the end, I got

```{r}
Transcript written on mf.log.
fmtutil [INFO]: /Users/danielapalleschi/Library/texlive/2019/texmf-var/web2c/metafont/mf.base installed.
fmtutil [INFO]: Disabled formats: 6
fmtutil [INFO]: Successfully rebuilt formats: 45
fmtutil [INFO]: Total formats: 51
fmtutil [INFO]: exiting with status 0
```

## Solution

In the end I went into my Time Machine (external harddrive) to a point right before this problem started happening, and restored the 'Macintosh HD/Library' folder, because presumably something happened to the TexLive file or something in there. That solved the problem.